---
title: MDS Notes
---

Notes from Conway & White Chapter 9 on multi-dimensional scaling.

We start out with a very simple matrix:

```{r results = 'asis'}
set.seed(851982) # To make sure results are consistent
ex.matrix <- matrix(sample(c(-1, 0, 1), 24, replace = TRUE),
                    nrow = 4,
                    ncol = 6)
row.names(ex.matrix) <- c('A', 'B', 'C', 'D')
colnames(ex.matrix) <- c('P1', 'P2', 'P3', 'P4', 'P5', 'P6')

pander::pander(ex.matrix)
```

We then multiply this matrix by its transpose. Why? To measure the distance between the items. The matrix multiplication is sort of a correlation measure. Important note: in this matrix, 0 implies no observation. So, the "variance" of an item measures the number of observations.

```{r results = 'asis'}
ex.mult <- ex.matrix %*% t(ex.matrix)
pander::pander(ex.mult)
```

We see that "B" has the greatest number of observations at four. A has only two.

```{r results = 'asis'}
ex.dist <- dist(ex.mult)
pander::pander(as.matrix(ex.dist))
```

Note that the distance beetween a point and itself is zero. Question: given a distance matrix, can we decompose it into observations?

Given the distance matrix, we now render that distance into two-dimensions. Conway & White make the clever example of trying to render four equidistant points in two dimensions. It can't be done.

```{r}
ex.mds <- cmdscale(ex.dist)
plot(ex.mds, type = 'n')
text(ex.mds, c('A', 'B', 'C', 'D'))
```

Groovy. A and D are close together. B and C are far apart from everything but themselves. Let's try that again, but with actual names and more complete data.

```{r results = 'asis'}
films <- c("The Godfather", "Mean Streets", "Jurassic Park", "The Avengers", "My Best Friends Wedding", "Despicable Me")
Me <- c(1, 1, -1, 1, -1, 1)
Inessa <- c(1, 1, 1, 1, 1, 1)
Joe <- c(1, 1, -1, -1, -1, -1)
Jane <- c(1, -1, 1, 1, -1, 1)
df <- rbind(Me, Inessa, Joe, Jane)
colnames(df) <- films

pander::pander(df)
```

```{r}
matCovar <- df %*% t(df)
pander::pander(matCovar)
```

```{r }
matDist <- dist(matCovar)
pander::pander(as.matrix(matDist))
```

```{r}
mdsFilms <- cmdscale(matDist)
plot(mdsFilms, type = 'n')
text(mdsFilms, rownames(df))
```

And now again, but let's check the distances between the films.

```{r}
dfFilm <- t(df)
matCovar <- dfFilm %*% df
matDist <- dist(matCovar)
mdsFilms <- cmdscale(matDist)
plot(mdsFilms, type = 'n')
text(mdsFilms, rownames(dfFilm))
```

Neat! The mafia films are poles apart from the other stuff. The Avengers and Despicable Me are right on top of one another. Not what I would expect, but consistent with this data.

One more time, but with some beer recipes taken from Charlie Papazian's Complete Joy of Homebrewing.

```{r}
dfBeer <- data.frame(BeerName = c("Sparrow Hawk", "Doctor Bock", "Toad Spit", "Potlach", "Palace Bitter")
                     , BeerStyle = c("Porter", "Starkbier", "Stout", "Starkbier", "Pale Ale")
                     , PageNumber = c(201, 202, 206, 233, 170)
                     , Yeast = c("Ale", "Ale", "Ale", "Lager", "Ale")
                     , Color = c(50, 28, 40, 28, 13)
                     , IBU = c(38, 26, 40, 33, 26)
                     , OG = c(1.058, 1.066, 1.060, 1.076, 1.040)
                     , FG = c(1.014, 1.014, 1.016, 1.018, 1.008))
```

```{r}
beerMat <- as.matrix(dfBeer[, c("Color", "IBU", "OG")])
row.names(beerMat) <- dfBeer$BeerName

beerMat <- beerMat %*% t(beerMat)
beerDist <- dist(beerMat)

beerMds <- cmdscale(beerDist)
row.names(beerMds) <- dfBeer$BeerStyle
```

```{r }
plot(beerMds)
plot(beerMds, type = "n")
text(beerMds, row.names(beerMds))
```

```{r}
data.dir <- file.path("09-MDS", "data", "roll_call")
data.files <- list.files(data.dir)

data.files

# Add all roll call vote data frames to a single list
rollcall.data <- lapply(data.files,
                        function(f)
                        {
                          read.dta(file.path(data.dir, f), convert.factors = FALSE)
                        })

dim(rollcall.data[[1]])

head(rollcall.data[[1]])
```

```{r}
rollcall.simplified <- function(df)
{
  no.pres <- subset(df, state < 99)
  
  for(i in 10:ncol(no.pres))
  {
    no.pres[,i] <- ifelse(no.pres[,i] > 6, 0, no.pres[,i])
    no.pres[,i] <- ifelse(no.pres[,i] > 0 & no.pres[,i] < 4, 1, no.pres[,i])
    no.pres[,i] <- ifelse(no.pres[,i] > 1, -1, no.pres[,i])
  }
  
  return(as.matrix(no.pres[,10:ncol(no.pres)]))
}

rollcall.simple <- lapply(rollcall.data, rollcall.simplified)
```

```{r}
rollcall.dist <- lapply(rollcall.simple, function(m) dist(m %*% t(m)))

rollcall.mds <- lapply(rollcall.dist,
                       function(d) as.data.frame((cmdscale(d, k = 2)) * -1))


congresses <- 101:111

for(i in 1:length(rollcall.mds))
{
  names(rollcall.mds[[i]]) <- c("x", "y")
  
  congress <- subset(rollcall.data[[i]], state < 99)
  
  congress.names <- sapply(as.character(congress$name),
                           function(n) strsplit(n, "[, ]")[[1]][1])
  
  rollcall.mds[[i]] <- transform(rollcall.mds[[i]],
                                 name = congress.names,
                                 party = as.factor(congress$party),
                                 congress = congresses[i])
}

head(rollcall.mds[[1]])
cong.110 <- rollcall.mds[[9]]

base.110 <- ggplot(cong.110, aes(x = x, y = y)) +
  scale_size(range = c(2,2), guide = 'none') +
  scale_alpha(guide = 'none') +
  theme_bw() +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank()) +
  ggtitle("Roll Call Vote MDS Clustering for 110th U.S. Senate") +
  xlab("") +
  ylab("") +
  scale_shape(name = "Party", breaks = c("100", "200", "328"),
              labels = c("Dem.", "Rep.", "Ind."), solid = FALSE) +
  scale_color_manual(name = "Party", values = c("100" = "blue", "200" = "red", "328"="grey"),
                     breaks = c("100", "200", "328"),
                     labels = c("Dem.", "Rep.", "Ind."))

base.110 + geom_text(aes(color = party, label = cong.110$name, size = 2))
```

```{r }
all.mds <- do.call(rbind, rollcall.mds)

all.plot <- ggplot(all.mds, aes(x = x, y = y)) +
  geom_point(aes(shape = party, size = 2, color = party)) +
  scale_size(range = c(2, 2), guide = 'none') +
  scale_alpha(guide = 'none') +
  theme_bw() +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank()) +
  scale_color_manual(name = "Party", values = c("100" = "blue", "200" = "red", "328"="grey"),
                     breaks = c("100", "200", "328"),
                     labels = c("Dem.", "Rep.", "Ind.")) + 
  ggtitle("Roll Call Vote MDS Clustering for U.S. Senate (101st - 111th Congress)") +
  xlab("") +
  ylab("") +
  scale_shape(name = "Party", breaks = c("100", "200", "328"), labels = c("Dem.", "Rep.", "Ind."), solid = FALSE) +
  facet_wrap(~ congress)

all.plot
```

```{r }
# This is the code omitted from the chapter.  This is used to create shnazy plots of everything!
for(i in 1:length(rollcall.mds))
{
  mds <- rollcall.mds[[i]]
  congress <- congresses[i]
  plot.title <- paste("Roll Call Vote MDS Clustering for ",
                      congress,
                      " U.S. Senate",
                      sep = "")
  
  # Build base plot
  mds.plot <- ggplot(mds, aes(x = x, y = y)) +
    scale_size(range = c(2, 2), guide = 'none') +
    scale_alpha(guide = 'none') +
    theme_bw() +
    theme(axis.ticks = element_blank(), 
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          panel.grid.major = element_blank()) +
    ggtitle(plot.title) +
    xlab("") +
    ylab("")
  
  # Build up point and text plots separately
  mds.point <- mds.plot + geom_point(aes(shape = party,
                                         alpha = 0.75,
                                         size = 2))
  mds.text <- mds.plot + geom_text(aes(color = party,
                                       alpha = 0.75,
                                       label = mds$name,
                                       size = 2))
  
  # Fix labels, shapes and colors
  if(length(levels(mds$party)) > 2)
  {
    mds.point <- mds.point + scale_shape(name = "Party",
                                         breaks = c("100", "200", "328"),
                                         labels = c("Dem.", "Rep.", "Ind."),
                                         solid = FALSE)
    mds.text <- mds.text + scale_color_manual(name = "Party",
                                              values = c("100" = "black",
                                                         "200" = "dimgray",
                                                         "328" = "gray"),
                                              breaks = c("100", "200", "328"),
                                              labels = c("Dem.", "Rep.", "Ind."))
  }
  else
  {
    mds.point <- mds.point + scale_shape(name = "Party",
                                         breaks = c("100", "200"),
                                         labels = c("Dem.", "Rep."),
                                         solid = FALSE)
    mds.text <- mds.text + scale_color_manual(name = "Party",
                                              values = c("100" = "black",
                                                         "200" = "dimgray"),
                                              breaks = c("100", "200"),
                                              labels = c("Dem.", "Rep."))
  }
  
}

```

